Nice guys or potential cooperators when keeping promises: an experimental and Bayesian account for two explanations

### 2019-07-29 article_english.Rmd
latex template
Translation

In the laboratory, these transgressions to trust have been explicitly studied in experiments where subjects have incentives to lie or not keep their promises. However, the vast majority of these studies in economics, psychology and neurosciences have been conducted in people who do NOT know each other [to name a few, @Gneezy2005; @Gneezy2013; @Baumgartner2009; @Baumgartner2013; @Charness2006; @Mazar2008; @Fischbacher2013a]. In social psychology some studies address deception in close interpersonal relationships, however, they use self-report measurements [@DePaulo1998; @DePaulo2004] or rather explore the development of deception detection skills in same-sex friends [@Anderson2002]. Despite the widespread presence of interactions between human beings belonging to the same group, it is not known how social closeness between subjects can affect keeping or breaking promises.

In this manuscript, we explore the effects of three partners with different levels of social closeness on keeping or breaking promises, as well as on cooperation in a trust game. The above, we analyze it in the light of the two main motivations that have been pointed out in the literature regarding keeping promises [@Baumgartner2009]:

@@

Subjects keep their promises by a combination between the _intrinsic_ and _instrumental_ motivations. They are predominantly _nice guys_ because they keep their word even with investors with no social closeness. However, its trustworthiness is far from perfect, since there is a small, but reliably greater than zero, proportion of betrayals committed towards this partner. It seems that if subjects cannot assure that their investor will be a _potential cooperator_, dishonesty may emerge in the form of broken promises. The theoretical predictions of the social norm _conditional cooperation_ frame the finding that social closeness decreases the probability of breaking the promise and increases cooperation. In the same way that suggests the correspondence between commitment expressed through promises and subsequent behavior.

>>>>

In the laboratory, these transgressions to trust have been explicitly studied in experiments where subjects have incentives to lie or not keep their promises. However, the vast majority of these studies in economics, psychology, and neurosciences have been conducted in people who do NOT know each other [to name a few, @Gneezy2005; @Gneezy2013; @Baumgartner2009; @Baumgartner2013; @Charness2006; @Mazar2008; @Fischbacher2013a]. In social psychology some studies address deception in close interpersonal relationships, however, they use self-report measurements [@DePaulo1998; @DePaulo2004] or rather explore the development of deception detection skills in same-sex friends [@Anderson2002]. Despite the widespread presence of interactions between human beings belonging to the same group, it is not known how social closeness between subjects can affect keeping or breaking promises.

In this manuscript, we explore the effects of three partners with different levels of social closeness on keeping or breaking promises, as well as on cooperation in a trust game. The above, we analyze it in light of the two main motivations that have been pointed out in the literature regarding keeping promises [@Baumgartner2009]:

@@

Subjects keep their promises by a combination of _intrinsic_ and _instrumental_ motivations. They are predominantly _nice guys_ because they keep their word even with investors with no social closeness. However, its trustworthiness is far from perfect, since there is a small, but reliably greater than zero, proportion of betrayals committed towards this partner. It seems that if subjects cannot assure that their investor will be a _potential cooperator_, dishonesty may emerge in the form of broken promises. The theoretical predictions of the social norm _conditional cooperation_ frame the finding that social closeness decreases the probability of breaking the promise and increases cooperation. In the same way that suggests the correspondence between commitment expressed through promises and subsequent behavior.

### 2019-07-29 article_english.tex
Translation

In the laboratory, these transgressions to trust have been explicitly
studied in experiments where subjects have incentives to lie or not keep
their promises. However, the vast majority of these studies in
economics, psychology and neurosciences have been conducted in people
who do NOT know each other (to name a few, Gneezy, 2005; Baumgartner,
Fischbacher, Feierabend, Lutz, \& Fehr, 2009; Baumgartner, Gianotti, \&
Knoch, 2013; Charness \& Dufwenberg, 2006; Fischbacher \& Föllmi-Heusi,
2013; Gneezy, Rockenbach, \& Serra-Garcia, 2013; Mazar, Amir, \& Ariely,
2008). In social psychology some studies address deception in close
interpersonal relationships, however, they use self-report measurements
(DePaulo, Ansfield, Kirkendol, \& Boden, 2004; DePaulo \& Kashy, 1998)
or rather explore the development of deception detection skills in
same-sex friends (Anderson, DePaulo, \& Ansfield, 2002). Despite the
widespread presence of interactions between human beings belonging to
the same group, it is not known how social closeness between subjects
can affect keeping or breaking promises.

In this manuscript, we explore the effects of three partners with
different levels of social closeness on keeping or breaking promises, as
well as on cooperation in a trust game. The above, we analyze it in the
light of the two main motivations that have been pointed out in the
literature regarding keeping promises (Baumgartner et al., 2009):

@@

Subjects keep their promises by a combination between the
\emph{intrinsic} and \emph{instrumental} motivations. They are
predominantly \emph{nice guys} because they keep their word even with
investors with no social closeness. However, its trustworthiness is far
from perfect, since there is a small, but reliably greater than zero,
proportion of betrayals committed towards this partner. It seems that if
subjects cannot assure that their investor will be a \emph{potential
cooperator}, dishonesty may emerge in the form of broken promises. The
theoretical predictions of the social norm \emph{conditional
cooperation} frame the finding that social closeness decreases the
probability of breaking the promise and increases cooperation. In the
same way that suggests the correspondence between commitment expressed
through promises and subsequent behavior.

>>>>

In the laboratory, these transgressions to trust have been explicitly
studied in experiments where subjects have incentives to lie or not keep
their promises. However, the vast majority of these studies in
economics, psychology, and neurosciences have been conducted in people
who do NOT know each other (to name a few, Gneezy, 2005; Baumgartner,
Fischbacher, Feierabend, Lutz, \& Fehr, 2009; Baumgartner, Gianotti, \&
Knoch, 2013; Charness \& Dufwenberg, 2006; Fischbacher \& Föllmi-Heusi,
2013; Gneezy, Rockenbach, \& Serra-Garcia, 2013; Mazar, Amir, \& Ariely,
2008). In social psychology some studies address deception in close
interpersonal relationships, however, they use self-report measurements
(DePaulo, Ansfield, Kirkendol, \& Boden, 2004; DePaulo \& Kashy, 1998)
or rather explore the development of deception detection skills in
same-sex friends (Anderson, DePaulo, \& Ansfield, 2002). Despite the
widespread presence of interactions between human beings belonging to
the same group, it is not known how social closeness between subjects
can affect keeping or breaking promises.

In this manuscript, we explore the effects of three partners with
different levels of social closeness on keeping or breaking promises, as
well as on cooperation in a trust game. The above, we analyze it in
light of the two main motivations that have been pointed out in the
literature regarding keeping promises (Baumgartner et al., 2009):

@@

Subjects keep their promises by a combination of \emph{intrinsic} and
\emph{instrumental} motivations. They are predominantly \emph{nice guys}
because they keep their word even with investors with no social
closeness. However, its trustworthiness is far from perfect, since there
is a small, but reliably greater than zero, proportion of betrayals
committed towards this partner. It seems that if subjects cannot assure
that their investor will be a \emph{potential cooperator}, dishonesty
may emerge in the form of broken promises. The theoretical predictions
of the social norm \emph{conditional cooperation} frame the finding that
social closeness decreases the probability of breaking the promise and
increases cooperation. In the same way that suggests the correspondence
between commitment expressed through promises and subsequent behavior.
